---
title: 'Example Data Cleaning: Reef Check'
author: "lkm"
date: "9/12/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning= FALSE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

library(tidyverse)
library(kableExtra)
  
```


 
## Read and Review Data
Goal= cleaned data frame with average percent coral cover per transect survey, with additional information needed for the master coral database.
```{r data}    

#read in original csv file

rc_carib_substrate <- read_csv("sample_data/Reef_Check_Caribbean_Substrate.csv")

# clean column names- lowercase, no spaces
rc_carib_substrate <- janitor::clean_names(rc_carib_substrate)

#Look at number of sites and countries in this dataset

length(unique(rc_carib_substrate$reef_name)) #612 unique reefs
length(unique(rc_carib_substrate$country)) #28 countries


#substrate_codes
unique(rc_carib_substrate$substrate_code)
length(unique(rc_carib_substrate$substrate_code)) #10 types

# Review a sample of data
rc_carib_substrate[1:10,] %>%
 kableExtra::kable(caption="Raw Reef Check Data") %>%
  kable_styling(bootstrap_options = "condensed", full_width=F)


```
</br></br>

# Cleaning Reef Check Data - Caribbean  
* Cleaning steps:  
    * A: Calculate percent cover of each substrate code PER SEGMENT  
        * 1- Group by reef id, year, data, segment code, substrate code  
        * 2- Create the "perc_seg" variable that is the % cover of each substrate code per segment  
      
    * B: Calculate percent cover of each substrate code PER SURVEY  
        * 1- Group by reef_id, year, date, and substrate code  
        * 2- Create variable "perc_cov", which is the average percent cover for the transect, based off the "perc_seg" column  ; Create n_samples by counting the number of segment codes, based on the grouping 
        
    * C: Clean columns to bind later with master_coral  
        * 1- Pick distinct columns to keep  
        * 2- Remove cols don't need  
        * 3- Combine cover and sd data for substrate code- need to do this because can't use spread on 2 cols  
        * 4- Spread data so coral and macro have their own column  
        * 5- Split the cover & sd data into 2 separate columns  
        * 6- Unite deg_min_sec data into 1 col for lat and lon- convert later  
        * 7- Rename and mutate some columns  
        * 8- Convert lat and lon into correct format  
        * 9- Final column type cleaning (numeric values, create 4 digit years)
```{r data_clean}


#A: percent cover of each substrate code PER SEGMENT (ie out of 40 points)
rc_carib_substrate <- rc_carib_substrate %>%
  group_by(reef_id, year, date, segment_code, substrate_code) %>%
  mutate(perc_seg= (total/40)*100)

  #Note, if try to do % by segment and by survey at same time, the values are the same because of the way group_by is done-- this is why i separated this process into 2 steps


#B: Percent cover of each substrate code PER SURVEY (out of 160 points, or average of all the segments for that survey)
  #Note: 8/20/18- Found that for some locations, there was more than 1 survey done on a particular day. To keep surveys separate, need to also group by "depth"
rc_carib_substrate <- rc_carib_substrate %>%
  group_by(reef_id, year, date, substrate_code, depth) %>%
  mutate(perc_cov= round(mean(perc_seg), 2),
         sd_cov= round(sd(perc_seg),2),
         n_samples = length(segment_code)) #this will tell me how many segments were completed!
  
#C: Clean columns to be in a good format to bind rows to master_coral dataframe

rc_carib_to_bind <- rc_carib_substrate %>% 
  #1 find distinct surveys- for reef check these are the unique variables
  distinct(reef_id, date, depth, substrate_code, .keep_all = TRUE) %>%
  
  #2 get rid of columns we no longer need
  select(-segment_code, -total, -state_province_island, -city_town, -perc_seg, - errors, -what_errors) %>%
  
  #3 filter out the 2 substrate codes of interest, hard coral (HC) & algae (NI)
  filter(substrate_code == "HC" | substrate_code == "NI") %>%
  
  #can't use spread on 2 cols, so combo % cov and sd, separated by _
  unite(temp, perc_cov, sd_cov, sep="_") %>%
  
  #4 use spread to put the % cover of each substrate code into its own column
  spread(substrate_code, temp) %>%
  
  #5 break apart HC and NI data back into cover and sd
  separate(HC, into= c("coral", "coral_std"), sep="_", remove=TRUE) %>%
  separate(NI, into= c("macro", "macro_std"), sep="_", remove= TRUE) %>%
  
  #6 Unite lat and lon col with deg_min_sec together
  unite(lat, latitude_degrees, latitude_minutes, latitude_seconds, sep =" ") %>%
  unite(lon, longitude_degrees, longitude_minutes, longitude_seconds, sep = " ") %>%
  
  #7 rename columns 
  rename(source_id = reef_id, depth_m=depth) %>%
  mutate(subregion = "caribbean",
         method= "line_transect",
         orig_source = "reef_check",
         data_source = "reef_check",
         t_length = 20) %>%
  separate(date, into= c("day", "month", "year"), sep="-", remove=TRUE)




#8: Converting the lat and lon from deg_min_sec to dec_deg formats. Use measurements package to do the conversion:

rc_carib_to_bind$lat <- measurements::conv_unit(rc_carib_to_bind$lat, "deg_min_sec", "dec_deg")

rc_carib_to_bind$lon <- measurements::conv_unit(rc_carib_to_bind$lon, "deg_min_sec", "dec_deg")

  #Make sure to put lat and lon as numeric
rc_carib_to_bind$lon <- as.numeric(rc_carib_to_bind$lon)
rc_carib_to_bind$lat <- as.numeric(rc_carib_to_bind$lat)

#9: Turn year into 4 digits (note: this data is from 2006-2017 so we can just add 20 to the beginning of each element)

rc_carib_to_bind$year <- paste(20, rc_carib_to_bind$year, sep="")
rc_carib_to_bind$day <- as.numeric(rc_carib_to_bind$day)
rc_carib_to_bind$year <- as.numeric(rc_carib_to_bind$year)

  #make coral data numeric
rc_carib_to_bind$coral <- as.numeric(rc_carib_to_bind$coral)
rc_carib_to_bind$coral_std <- as.numeric(rc_carib_to_bind$coral_std)
rc_carib_to_bind$macro <- as.numeric(rc_carib_to_bind$macro)
rc_carib_to_bind$macro_std <- as.numeric(rc_carib_to_bind$macro_std)


```
</br></br>

# Check Data
```{r data_check}
# Preview a sample of data:

rc_carib_to_bind[1:10,] %>%
  arrange(reef_name) %>%
  kableExtra::kable(caption="Coral and Macroalgae % Cover from Caribbean Reef Check Surveys") %>%
  kable_styling(bootstrap_options = "condensed", full_width=F)
```

