---
title: "Visualizing Mixed-effects Models"
author: "Laura Mudge"
date: "`r format(Sys.Date())`"
output:
  html_document:
    keep_md: TRUE
editor_options: 
  chunk_output_type: console
---
# What's in this document:
Some neat things I've learned about when handling mixed-effects models. The focus of these first few examples is how to *visualize* results of mixed-effects models.  

# Setup
Data = use the "mixedeff_herbivore.csv" file in the sample_data folder. This is a dataset used to explore the influences of herbivore populations on coral cover.  

```{r setup, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) #for all data wrangling
library(cowplot) #for manuscript ready figures
library(lme4) #for lmer & glmer models
library(sjPlot) #for plotting lmer and glmer mods
library(sjmisc) 
library(effects)
library(sjstats) #use for r2 functions


me_data <- read_csv("C:/github/sample_code/sample_data/mixedeff_herbivore.csv")

```

# Create a basic mixed-effects model:
I'm not going to walk through the steps to building models (at least not yet), but rather just show an example of a model with coral cover as the response variable (elkhorn_LAI), herbivore populations & depth as fixed effects (c.urchinden, c.fishmass, c.maxD), and survey site as a random effect (site).  
.  
Note: due to the difference in scale of how the herbivore populations are measured, we are using the centered & scaled values- otherwise models won't converge. We are also use the log of the response variable. I am subsetting the data based on this specific study. Here we are only using data for when LAI_nonzero==1.
```{r model}

#use the lme4 package

mod <- lme4::lmer(log(elkhorn_LAI) ~ c.urchinden + c.fishmass +c.maxD + (1|site), REML= FALSE, data= subset(me_data, LAI_nonzero ==1))
summary(mod)
```

# Make a plot of the effect sizes:
This would definitely be useful if you have a lot of fixed effects!

## Unformatted plot of effect sizes
```{r plotmod}

sjPlot::plot_model(mod)

```

## Formatted plot of effect sizes:
Let's change the axis labels & title. Type ?plot_model into your console to see details of ALL the features you can adjust.
```{r plotmod2}

# Notes: axis labels should be in order from bottom to top. 
# To see the values of the effect size and p-value, set show.values and show.p= TRUE. Pvalues will only be shown if the effect size values are too

sjPlot::plot_model(mod, 
                   axis.labels=c("Urchin", "Depth", "Fish"),
                   show.values=TRUE, show.p=TRUE,
                   title="Effect of Herbivores on Coral Cover")

```

# Table output of model results:
There's a neat feature of sjPlot that also creates nice tables of model summary outputs. This will give you the predictor variables included, their estimates, confidence intervals, p-values for estimates, and random effects information. Type ?tab_model in your console to see all the features you can adjust.

### Unformatted table
```{r mod_table}

sjPlot:: tab_model(mod)

```

### Formatted table
```{r mod_table2}
# Notes: predictor labels (pred.labels) should be listed from top to bottom; dv.labels= the name of the response variable that will be at the top of the table.
sjPlot::tab_model(mod, 
                  show.re.var= TRUE, 
                  pred.labels =c("(Intercept)", "Urchins", "Fish", "Depth"),
                  dv.labels= "Effects of Herbivores on Coral Cover")
```

# Plot model estimates WITH data
Using the 'effects' and 'ggplot2' packages, we can plot the model estimates on top of the actual data! We do this for one variable at a time. Note: the urchin data was scaled & centered for use in the model, so we are plotting the scaled and centered data values NOT the raw data (ie urchin density) 

### Step 1: Save the effect size estimates into a data.frame
```{r}
# Use the effects package --> effect function. term= the fixed effect you want to get data on, mod= name of your model.

effects_urchin <- effects::effect(term= "c.urchinden", mod= mod)
summary(effects_urchin) #output of what the values are

# Save the effects values as a df:
x_urch <- as.data.frame(effects_urchin)

```

### Step 2: Use the effects value df (created above) to plot the estimates
You can break this up into separate steps if you wish to save a base plot (of your fixed effect & response var data only). Note: for the plot, I am subsetting the data based on this specific study. Here we are only using data for when LAI_nonzero==1.
```{r}

#Basic steps:
  #1 Create empty plot

  #2 Add geom_points() from the DATA: urchin data on the x axis (independent va= c.urchinden) and coral data on the y-axis (response var= elkhorn_LAI)

  #3 Add geom_point for the MODEL estimates (data= x_urchi here, this is the dataset you created in the above chunk). We will change the color so they are distinguishable from the data

  #4 Add geom_line for the MODEL estimates. Change the color to match the estimate points (ie whatever color you chose for step3)

  #5 Add geom_ribbon that has the CI limits for the model estimates

  #6 Edit the labels as you see fit!

#1
urchin_plot <- ggplot() + 
  #2
geom_point(data=subset(me_data, LAI_nonzero==1), aes(c.urchinden, log(elkhorn_LAI))) + 
  #3
  geom_point(data=x_urch, aes(x=c.urchinden, y=fit), color="blue") +
  #4
  geom_line(data=x_urch, aes(x=c.urchinden, y=fit), color="blue") +
  #5
  geom_ribbon(data= x_urch, aes(x=c.urchinden, ymin=lower, ymax=upper), alpha= 0.3, fill="blue") +
  #6
  labs(x="Urchins (centered & scaled)", y="Coral Cover")

urchin_plot

```

